Proof nets for first-order additive linear logic

### 2018-08-07 ALL1.tex
further referee-induced changes

We conclude this section by establishing that sequentialization and de-sequentialization for witness nets are inverses.

@@

A \defn{slice} is the fraction of a proof that depends on a given choice of one branch (or projection) on each product formula $A\*B$. Important to additive proof theory is that many operations can be performed on a per-slice basis, such as normalization, or proof net correctness. We will here use slices for the latter purpose.
%
As in the propositional case \cite{Hughes-vanGlabbeek-2005}, we define a \defn{slice} of a sequent $\seq AB$ as a set of potential links, of which exactly one must be realized in a proof net $\net{\lambda_\Sigma}AB$. We extend the propositional criterion in two ways:

\begin{description}
	\item[Expansion]
When defining slices, we interpret an existential quantification $\exists x.A$ as a sum over all witnesses $t_i$ to $x$ that occur in the pre-net, $A[t_1/x]\+\dots\+A[t_n/x]$. This captures the non-permutability of a product rule over distinct instantiations, as below left.
	\item[Dependency]
We define a \defn{dependency} relation between a universal quantification $\forall x.A$ and an instantiation $B[t/y]$ of $\exists x.B$ where the \defn{eigenvariable} $x$ occurs free in $t$ (a standard approach to first-order quantification \cite{Miller-1987,Girard-1996,Heijltjes-2010,Mimram-2011}). For each link, we require the dependency to be \emph{acyclic}, which amounts to \emph{slice-wise} first-order correctness. It captures the non-permutability of universal and existential sequent rules due to the \emph{eigenvariable condition} of the former, as below right.
	\item[]
\[
\vc{
	\infer[\Qrr*]{\Seq{\exists x.A}{B\*C}}{
	 \infer[\Qrr?s]{\Seq{\exists x.A}B}{\Seq{A[s/x]}B}
	 &
	 \infer[\Qrr?t]{\Seq{\exists x.A}C}{\Seq{A[t/x]}C}
   }}
\qquad\qquad
\vc{
	\infer[\Qrr!]{\Seq{\forall x.A}{\exists y.B}}{
	 \infer[\Qrr?t]{\Seq A{\exists y.B}}{\Seq AB[t/y]}
   }}
   \quad\text{where }x\in\fv(t)
\]
\end{description}

\noindent
Formalizing this, for a witness linking $\lambda_\Sigma$ and variable $x$, write $\Sigma(x)\subseteq\terms$ for the \defn{witness set} of $x$, which collects the terms assigned to $x$ by the witness maps in $\Sigma$ for every link in $\lambda$:
\[
	\Sigma(x) = \{~\sigma(x)~\mid~\link CD\in\lambda_\Sigma~\}~.
\]

% ..................................................
\begin{definition}[Slice]
Given a witness linking $\lambda_\Sigma$, a \defn{slice} $S$ of a formula $A$ and a witness map $\sigma$ is a set of pairs $(A',\sigma')$, where $A'\leq A$ and $\sigma'\supseteq\sigma$, given by $S=\{(A,\sigma)\}\cup S'$ where:
%
\begin{itemize}
	\item
\makebox[65pt][l]{If $A = a$}			then $S'=\varnothing$.
	\item
\makebox[65pt][l]{If $A = B\+C$} 		then $S'=S_B\uplus S_C$ with $S_B$ a slice of $B$ and $\sigma$, and $S_C$ one of $C$ and $\sigma$.
	\item
\makebox[65pt][l]{If $A = B\*C$}		then $S'$ is a slice of $B$ and $\sigma$ or a slice of $C$ and $\sigma$.
	\item
\makebox[65pt][l]{If $A = \exists x.B$}	then $S'=\uplus_{t\,\in\,\Sigma(x)}S_t$ where each $S_t$ is a slice of $B$ and $\sigma[t/x]$.
	\item
\makebox[65pt][l]{If $A = \forall x.B$}	then $S'$ is a slice of $B$ and $\sigma$.
\end{itemize}
%
\noindent
A \defn{slice} of a sequent $\seq AB$ is a set of links
\[
	\{~\link[\sigma\cup\tau]CD~\mid~(C,\sigma)\in S_A~,~(D,\tau)\in S_B~\}
\]
where $S_A$ is a slice of $A$ and $\varnothing$, and $S_B$ a slice of $B$ and $\varnothing$.  A \defn{slice} of a witness pre-net $\net{\lambda_\Sigma} AB$ is the intersection $\lambda_\Sigma\cap S$ of $\lambda_\Sigma$ with a slice $S$ of $\seq AB$.
\end{definition}
% ..................................................

@@

\begin{definition}
In a pre-net $\net{\lambda_\Sigma}AB$, let the \defn{column} of a link $\link CD$ be the set of pairs 
\[
	\{~(X,\sigma\res{\ex[A]X})~\mid~C\sub X\sub A~\}\cup\{~(Y,\sigma\res{\ex[B]Y})~\mid~D\sub Y\sub B~\}~,
\]
with a \defn{dependency} order $(\dep)$: $(X,\rho)\dep(Y,\tau)$ if $X\leq Y$ or $Y$ occurs as $\forall x.Y$ and $x\in\rho$.
\end{definition}

@@

% ..................................................
\begin{figure}[!t]
\[
\begin{array}{ccc}
	\vc{
	\infer[\!\!\scriptstyle{\mathrm{cut}}]{\Seq AC}{
	 \infer[\!\!\scriptstyle{\qrr*}]{\Seq A{B_1\*B_2}\vphantom{\dual B}}{\Prf{\pi_1}A{B_1} & \Prf{\pi_2}A{B_2}}
	 &&
	 \infer[\!\!\scriptstyle{\qrr+i}]{\Seq {\dual B_1\+\dual B_2}C}{\Prf\phi{B_i}C}
	}}
	&\Rightarrow&
	\vc{\infer[\!\!\scriptstyle{\mathrm{cut}}]{\Seq AC}{\Prf{\pi_i}A{B_i} && \Prf\phi{\dual B_i}C}}
\\ \\
	\vc{
	\infer[\!\!\scriptstyle{\mathrm{cut}}]{\Seq AC}{
	 \infer[\!\!\scriptstyle{\qrr?t}]{\Seq A{\exists x.B}}{\Prf\pi A{B[t/x]}}
	 &&
	 \infer[\!\!\scriptstyle{\qrr!}]{\Seq {\forall x.\dual B}C}{\Prf\phi{\dual B}C}
	}}
	&\Rightarrow&
	\vc{\infer[\!\!\scriptstyle{\mathrm{cut}}]{\Seq AC}{\Prf\pi A{B[t/x]} && \Prf{\phi[t/x]}{\dual B[t/x]}C}}
\end{array}
\]
\caption{\all1 cut-elimination steps}
\label{fig:cut-elim}
\end{figure}
% ..................................................

@@

%
A link $\link[]ab$ on two atomic formulas is an \defn{axiom} link if there exists a witness map $\sigma$ such that $a\sigma=\dual b\sigma$. To an axiom link $\link[]ab$ over $\seq AB$ we assign an \defn{initial witness map}, which is the least witness map $\sigma$ over the domain $\ex[\!A]a\cup\ex[\!B]b$ such that $a\sigma=\dual b\sigma$. In other words, $\sigma$ is the most general unifier of $a$ and $\dual b$, over the given domain, written $\init ab$. For an axiom linking $\lambda$ over $\seq AB$ the \defn{initial witness pre-net} $\net{\lambda_\star}AB$ is given by
\[
	\lambda_\star = \{~\link ab~\mid~\link[]ab\in\lambda~,~\sigma=\init ab~\}~.
\]
An initial witness pre-net has \emph{exact coverage}, while coalescence will give \emph{local eigenvariables}.

@@

In the above definition, note that $\lambda_\star^\star=(\lambda_\star)^*$ is the initial proof labelling of $\lambda_\star$, which assigns an axiom rule to each axiom link. For a minimal example, see the introduction. Observe also that unifying coalescence includes strict coalescence, $(\scoal)\subseteq(\ucoal)$. The following two lemmata relate sequentialization for witness nets and unification nets.

@@

Our slicing condition is closely related to a number of approaches to first-order classical logic. A formula $\exists x.A$ is interpreted as the sum (or classically, the \emph{disjunction}) over a fixed number of instantiations $A[t_1/x]\+\dots\+A[t_n/x]$. This can be traced back to Herbrand's Theorem \cite{Herbrand-1930}: $\exists x.A$ is equivalent to the infinite sum over $A[t/x]$ for all terms $t$ in the language, but for any given proof a finite set of terms suffices. \emph{Expansion tree proofs} \cite{Miller-1987,Heijltjes-2010} are a graphical proof formalism based on this idea.

In our slicing condition, the witnesses $t_1,\dots,t_n$ for the expansion $A[t_1/x]\+\dots\+A[t_n/x]$ are collected from the substitutions recorded at the links of the proof net. Expansion trees, in contrast, record the expansion and its terms explicitly. This could be an interesting alternative approach to additive proof nets, which we may explore in future work. It is expected, though, that this would sacrifice \emph{generality} and \emph{direct composition}.

@@

\begin{proof}
For a strict coalescence step $L\scoal R$, we will show that the witness pre-net $L$ is correct if and only if $R$ is. Let $L=\net{\lambda_\Sigma}AB$ and $R=\net{\kappa_\Theta}AB$. In each case, exact coverage is immediately preserved and reflected. For slice-correctness, we will demonstrate that the left-hand side and right-hand side of each rule belong to the same slice of $\seq AB$, or in the case of $\srr?$, naturally corresponding slices. For dependency-correctness, we will briefly show how acyclicity of the columns of the involved links is preserved.

@@

	\item
$\link CD\scoal\link[\sigma\minus x]C{\exists x.D}$
\\
The witness sets $\Sigma(x)$ and $\Theta(x)$ for $L$ and $R$ need not be the same, since $L$ has $\sigma$ where $R$ has $\sigma\minus x$. Let $\Sigma(x)=\{t,t_1,\dots,t_n\}$ and $\Theta(x)=\{t_1,\dots,t_n\}$, where $\sigma(x)=t$. For every slice $S_B$ of $B$ over $\Sigma$ there is a corresponding slice $S'_B$ over $\Theta$, both (or neither) containing $(\exists x.D,\tau)$ and $(D,\tau[t_i/x])$ for $i\leq n$, but the former in addition having $(D,\tau[t/x])$. Then $S\cap \lambda_\Sigma$ is the singleton $\{\link CD\}$ if and only if $S\cap\kappa_\Theta$ is $\{\link[\sigma\minus x]C{\exists x.D}\}$, where it should be observed that if $t$ is also a witness to $x$ in some other slice of $R$, then $t\in\Theta(x)$ and in fact $S=S'$.

>>>>

We conclude this section by establishing that sequentialization and de-sequentialization for witness nets are inverses, and that witness nets are canonical.

@@

A \defn{slice} is the fraction of a proof that depends on a given choice of one branch (or projection) on each product formula $A\*B$. Important to additive proof theory is that many operations can be performed on a per-slice basis, such as normalization, or proof net correctness. We will here use slices for the latter purpose. As in the propositional case \cite{Hughes-vanGlabbeek-2005}, we define a \defn{slice} of a sequent $\seq AB$ as a set of potential links, of which exactly one must be realized in a proof net $\net{\lambda_\Sigma}AB$. We extend the propositional criterion in two ways:

\begin{description}
	\item[Expansion]
When defining slices, we interpret an existential quantification $\exists x.A$ as a sum over all witnesses $t_i$ to $x$ that occur in the pre-net, $A[t_1/x]\+\dots\+A[t_n/x]$. This captures the non-permutability of a product rule over distinct instantiations, as below left. (Technically, a slice of $\exists x.A$ will correspond to an infinite sum over $A[t/x]$ for every term $t$, but only the actually occurring terms $t_i$ will ever be relevant.)

	\item[Dependency]
We define a \defn{dependency} relation between a universal quantification $\forall x.A$ and an instantiation $B[t/y]$ of $\exists y.B$ where the eigenvariable $x$ occurs free in $t$ (a standard approach to first-order quantification \cite{Miller-1987,Girard-1996,Heijltjes-2010,Mimram-2011}). For each link, we will require this dependency relation to be \emph{acyclic}, which amounts to \emph{slice-wise} first-order correctness. It captures the non-permutability of universal and existential sequent rules due to the \emph{eigenvariable condition} of the former, as below right.
	\item[]
\[
\vc{
	\infer[\Qrr*]{\Seq{\exists x.A}{B\*C}}{
	 \infer[\Qrr?s]{\Seq{\exists x.A}B}{\Seq{A[s/x]}B}
	 &
	 \infer[\Qrr?t]{\Seq{\exists x.A}C}{\Seq{A[t/x]}C}
   }}
\qquad\qquad
\vc{
	\infer[\Qrr!]{\Seq{\forall x.A}{\exists y.B}}{
	 \infer[\Qrr?t]{\Seq A{\exists y.B}}{\Seq AB[t/y]}
   }}
   \quad\text{where }x\in\fv(t)
\]
\end{description}

%\noindent
%Formalizing this, for a witness linking $\lambda_\Sigma$ and variable $x$, write $\Sigma(x)\subseteq\terms$ for the \defn{witness set} of $x$, which collects the terms assigned to $x$ by the witness maps in $\Sigma$ for every link in $\lambda$:
%\[
%	\Sigma(x) = \{~\sigma(x)~\mid~\link CD\in\lambda_\Sigma~\}~.
%\]

% ..................................................
\begin{definition}[Slice]
A \defn{slice} $S$ of a formula $A$ and a witness map $\sigma$ is a set of pairs $(A',\sigma')$, where $A'\leq A$ and $\sigma'\supseteq\sigma$, given by $S=\{(A,\sigma)\}\cup S'$ where:
%
\begin{itemize}
	\item
\makebox[65pt][l]{If $A = a$}			then $S'=\varnothing$.
	\item
\makebox[65pt][l]{If $A = B\+C$} 		then $S'=S_B\uplus S_C$ with $S_B$ a slice of $B$ and $\sigma$, and $S_C$ one of $C$ and $\sigma$.
	\item
\makebox[65pt][l]{If $A = B\*C$}		then $S'$ is a slice of $B$ and $\sigma$ or a slice of $C$ and $\sigma$.
	\item
\makebox[65pt][l]{If $A = \exists x.B$}	then $S'=\uplus_{t\,\in\,\terms}S_t$ where each $S_t$ is a slice of $B$ and $\sigma[t/x]$.
	\item
\makebox[65pt][l]{If $A = \forall x.B$}	then $S'$ is a slice of $B$ and $\sigma$.
\end{itemize}
%
\noindent
A \defn{slice} of a sequent $\seq AB$ is a set of links
\[
	\{~\link[\sigma\cup\tau]CD~\mid~(C,\sigma)\in S_A~,~(D,\tau)\in S_B~\}
\]
where $S_A$ is a slice of $A$ and $\varnothing$, and $S_B$ a slice of $B$ and $\varnothing$.  A \defn{slice} of a witness pre-net $\net{\lambda_\Sigma} AB$ is the intersection $\lambda_\Sigma\cap S$ of $\lambda_\Sigma$ with a slice $S$ of $\seq AB$.
\end{definition}
% ..................................................

@@

\begin{definition}
In a pre-net $\net{\lambda_\Sigma}AB$, let the \defn{column} of a link $\link CD$ be the set of pairs 
\[
	\{~(X,\sigma\res{\ex[A]X})~\mid~C\sub X\sub A~\}\cup\{~(Y,\sigma\res{\ex[B]Y})~\mid~D\sub Y\sub B~\}~,
\]
with a \defn{dependency} relation $(\dep)$: $(X,\rho)\dep(Y,\tau)$ if $X\leq Y$ or $Y$ occurs as $\forall x.Y$ and $x\in\rho$.
\end{definition}

@@

% ..................................................
\begin{figure}[!t]
\[
\begin{array}{ccc}
	\vc{
	\infer[\!\!\scriptstyle{\mathrm{cut}}]{\Seq AC}{
	 \infer[\!\!\scriptstyle{\qrr*}]{\Seq A{B_1\*B_2}\vphantom{\dual B}}{\Prf{\pi_1}A{B_1} & \Prf{\pi_2}A{B_2}}
	 &&
	 \infer[\!\!\scriptstyle{\qrr+i}]{\Seq {\dual B_1\+\dual B_2}C}{\Prf\phi{\dual B_i}C}
	}}
	&\Rightarrow&
	\vc{\infer[\!\!\scriptstyle{\mathrm{cut}}]{\Seq AC}{\Prf{\pi_i}A{B_i} && \Prf\phi{\dual B_i}C}}
\\ \\
	\vc{
	\infer[\!\!\scriptstyle{\mathrm{cut}}]{\Seq AC}{
	 \infer[\!\!\scriptstyle{\qrr?t}]{\Seq A{\exists x.B}}{\Prf\pi A{B[t/x]}}
	 &&
	 \infer[\!\!\scriptstyle{\qrr!}]{\Seq {\forall x.\dual B}C}{\Prf\phi{\dual B}C}
	}}
	&\Rightarrow&
	\vc{\infer[\!\!\scriptstyle{\mathrm{cut}}]{\Seq AC}{\Prf\pi A{B[t/x]} && \Prf{\phi[t/x]}{\dual B[t/x]}C}}
\end{array}
\]
\caption{\all1 cut-elimination steps}
\label{fig:cut-elim}
\end{figure}
% ..................................................

@@

%
A link $\link[]ab$ on two atomic formulas is an \defn{axiom} link if there exists a witness map $\sigma$ such that $\dual a\sigma=b\sigma$. To an axiom link $\link[]ab$ over $\seq AB$ we assign an \defn{initial witness map}, which is the least witness map $\sigma$ over the domain $\ex[\!A]a\cup\ex[\!B]b$ such that $\dual a\sigma=b\sigma$. In other words, $\sigma$ is the most general unifier of $\dual a$ and $b$, over the given domain, written $\init{\dual a}b$. For an axiom linking $\lambda$ over $\seq AB$ the \defn{initial witness pre-net} $\net{\lambda_\star}AB$ is given by
\[
	\lambda_\star = \{~\link ab~\mid~\link[]ab\in\lambda~,~\sigma=\init{\dual a}b~\}~.
\]
An initial witness pre-net has \emph{exact coverage}, while coalescence will give \emph{local eigenvariables}.

@@

In the above definition, note that $\lambda_\star^\star=(\lambda_\star)^\star$ is the initial proof labelling of $\lambda_\star$, which assigns an axiom rule to each axiom link. For a minimal example, see the introduction. Observe also that unifying coalescence includes strict coalescence, $(\scoal)\subseteq(\ucoal)$. The following two lemmata relate sequentialization for witness nets and unification nets.

@@

Our slicing condition is loosely related to a number of approaches to first-order classical logic. A formula $\exists x.A$ is interpreted as the sum (or classically, the \emph{disjunction}) over a fixed number of instantiations $A[t_1/x]\+\dots\+A[t_n/x]$. This can be traced back to Herbrand's Theorem \cite{Herbrand-1930}: $\exists x.A$ is equivalent to the infinite sum over $A[t/x]$ for all terms $t$ in the language, but for any given proof a finite set of terms suffices. \emph{Expansion tree proofs} \cite{Miller-1987,Heijltjes-2010} are a graphical proof formalism based on this idea.

In our slicing condition, the interpretation of $\exists x.A$ is an infinite sum of which only a finite part $A[t_1/x]\+\dots\+A[t_n/x]$ is relevant, over the witnesses $t_1,\dots,t_n$ actually assigned to $x$ in the proof net. An interesting alternative approach to additive proof nets, which we may explore in future work, could be to take the expansion of $\exists x.A$ to $A[t_1/x]\+\dots\+A[t_n/x]$ as primary, and record it explicitly in the syntax, as expansion tree proofs do for classical logic. It is expected, however, that this would sacrifice \emph{generality} and \emph{direct composition}.

@@

\begin{proof}
For a strict coalescence step $L\scoal R$, we will show that the witness pre-net $L$ is correct if and only if $R$ is. Let $L=\net{\lambda_\Sigma}AB$ and $R=\net{\kappa_\Theta}AB$. In each case, exact coverage and local eigenvariables are immediately preserved and reflected. For slice-correctness, we will demonstrate that the left-hand side and right-hand side of each rule belong to the same slice of $\seq AB$, or in the case of $\srr?$, naturally corresponding slices. For dependency-correctness, we will briefly show how acyclicity of the columns of the involved links is preserved.

@@

	\item
$\link CD\scoal\link[\sigma\minus x]C{\exists x.D}$
\\
A slice $S$ of $\seq AB$ contains $\link[\tau] C{\exists x.D}$ if and only if it contains all links $\link[\tau{[t/x]}]CD$ for any term $t$. Letting $\tau[t/x]=\sigma$, then $S\cap \lambda_\Sigma$ is the singleton $\{\link CD\}$ if and only if $S\cap\kappa_\Theta$ is $\{\link[\sigma\minus x]C{\exists x.D}\}$.
%
%The witnesses sets $\Sigma(x)$ and $\Theta(x)$ for $L$ and $R$ need not be the same, since $L$ has $\sigma$ where $R$ has $\sigma\minus x$. Let $\Sigma(x)=\{t,t_1,\dots,t_n\}$ and $\Theta(x)=\{t_1,\dots,t_n\}$, where $\sigma(x)=t$. For every slice $S_B$ of $B$ over $\Sigma$ there is a corresponding slice $S'_B$ over $\Theta$, both (or neither) containing $(\exists x.D,\tau)$ and $(D,\tau[t_i/x])$ for $i\leq n$, but the former in addition having $(D,\tau[t/x])$. Then $S\cap \lambda_\Sigma$ is the singleton $\{\link CD\}$ if and only if $S\cap\kappa_\Theta$ is $\{\link[\sigma\minus x]C{\exists x.D}\}$, where it should be observed that if $t$ is also a witness to $x$ in some other slice of $R$, then $t\in\Theta(x)$ and in fact $S=S'$.
